
---
title: "PS0: Intro R and RMarkdown"
author: "Joseph Rebello"
output: 
  html_document:
    theme: readable
    highlight: pygments

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r r_setup, echo=FALSE, warnings = FALSE, message = FALSE}

library(wooldridge)
library(tidyverse)

setwd("C:/Users/josep/Dropbox/TCHNG/spring20")


```


## Hi

The goal of this assignment, due on the first day of class, is to get our technology set up and climb some of the initial learning curve. This is not a comp sci course in any sense, but we will be using our computers constantly. Nothing has to be easy at first, but we don't want to be dealing with a problem installing essential software the night before a substantial problem set. This problem set assures our technology is working by day 1 and allows us (I'm here to help) to solve any technical issues before we get into real content. I also hope it gives you some time to play around on your computer. If the only time you use class technology is when directly instructed to, you will never be comfortable with it.

If you have an R background, you can just update this code and submit as directed. If not, you are encouraged to view my walkthrough video. I leave some exercises up to you but I'll walk you through most of it with significant explanation. If you're at a place where you want some assistance but have enough background to make some of my explanations excessive, I have heard the adjust-speed option is good. Because I am encouraging you to follow along me you should not take the difficulty (relative to where you may be right now) and vagueness of the questions as typical of future problem sets. I've written this as if you are following along with me.

## Stuff We'll Use

The first thing we need to do is get our computer ready. First, we need to download and install R and RStudio. Once we have RStudio running we will then install the packages rmarkdown and knitr. While we're at it we can install the wooldridge and tidyverse packages.


## Production and Distribution

1. I will distribute your problem sets in at least two files. A html version will be for viewing your problem set. There will be questions and R code fragments. The second file will be the .Rmd file Knitr used to produce the pdf/html document. Other files may be shared as well, including .bib files for bibliographic information and .csv's for datasets. In general, any file I post in your homework folder should be downloaded. Anything other than the html meant for your reviewing will be essential for producing your submission.

2. You will download the .Rmd file and begin editing it. Start with changing the name. If a question asks for a written response, type your answer in the appropriate space. If the question asks you to modify the code, go ahead and modify the code. If you are instructed to do something by hand on paper, do that too and remember to bring it to class.

3. You will create a pdf or html file once you're done answering, modifying, and completing. You then submit the pdf/html and the .Rmd file via blackboard. After submitting, think about how nice this is. Not only do I get to see your answers, I get to see exactly how you found/created your answers. I really do need your .Rmd file. Correct answers without the code to reproduce them will not get you credit.


## Let's Go! Edit Text and Settings

1. If you're ready, please change the following blockquote.

> I'M NOT READY!!!

2. Check out your YAML. The YAML is the text between the triple dashes at the top of your document. It is meant to set properties of the document, like author and date. If we were creating a document in Word these properties would be hidden away somewhere in some menu. Does the YAML claim Joseph Rebello is the author of this document?  If so, go ahead and change it. You are the author now.

3. Update your working directory in the r_setup code chunk. This should be the same directory "Nightingale.csv"" is in.

4. Describe a topic that you'd like to research that there might be data for. In one legit paragraph describe the general topic, specific research question, and type of data you would use. 

I would like to research...


## Updating Code: Basic R Objects and Output

Throughout the rest of this assignment you will be updating R code. I hope this updating process helps begin your familiarization with R, but if things seem strange do not worry. Remember, we want to make sure things work (even if you're not 100% sure yet how) and being the process of familiarization.


```{r variable_creation, eval = FALSE}

# create some character variables using "quotes"

first_name <- 
last_name <- 
favorite_book <- 
favorite_movie <- 
undergrad_major <- 

# create some numeric variables
  
MA_semesters_completed <- 
micro_interest <- 
macro_interest <- 
pe_interest <- 
methods_interest <- 
  
# create some logical variables: TRUE or FALSE
  
undergrad_stats <- 
undergrad_metics <- 
grad_metrics <- 

# create a character vector use c()
  
favorite_foods <- 

# create a numeric vector by combining your interest variables from above
  
interests <- 

# create a logical vector: each element should be true if the corresponding element in interests is greater than 5.

interested <- 


```

Ok, now we can create some output. 

```{r ratings_barplot, eval=FALSE}

# Here we will create a barplot illustrating our interests in micro, macro, pe, metrics. We can use help('barplot') to see how to construct one.

barplot(interests)


```

Now we'll do a little work with dataframes. First, we'll simulate some data, and then import from a csv ("Nightingale.csv") and a package.

In the following chunk, complete the code so that Y is a linear function of X1 and X2 with an intercept of zero. Combine them in a dataframe. Output a table of summary statistics. Create a scatterplot of Y and X1.


```{r eval = FALSE}

# Create a variable n for our sample size. Choose any number above 50.

n <- 

# Use rnorm() to generate n random variables from a normal distribution with a mean and standard deviation of your choice. If you don't know what these terms mean, that is fine. Assign these values to X1. Repeat with a different mean and standard deviation for X2.

X1 <- 
X2 <- 

# Generate Y

Y <- 

# create dataframe

random_df <- 


# use summary() 

summary()

# use plot() to produce a scatterplot of Y and X1

plot()


```



When you run the previous chunk you will see a very ugly summary of our dataset. Let's pretty it up a tiny bit and learn some new syntax.


```{r eval = FALSE}

# use knitr::kable() to make everything prettier


```


Ok, let's do some work with actual data. We won't do too much with each, but you are encouraged to play around in a seperate R script. If you produce anything interesting feel free to share it in this document. First, we'll input data from a csv. In the second example we'll load from a data package. Later on in the semester you'll be exposed to other means of getting data into R.

You should have downloaded the "Nightingale.csv" file. In the following chunk, we'll import that data and use base R and tidyverse (ggplot) to make some line charts.


```{r eval = FALSE}

# import the nightingale data using the read.csv() function, name it nightingale

nightingale <- 

nightingale$Date <- 

# create a vector of 3 colors for our chart

colors <- 

# lets plot using plot(), lines() and legend() functions.




```


Ok, now we'll load some data via a package and make some scatter plots. The dataset is from the 1976 CPS so if any numbers seem odd, that might be why.

```{r eval=FALSE}

# load the wage1 data from wooldridge using... data().


data()


# create factor columns gender and marital based on female and married with levels 0,1 and lables "Male" "Female" and "Not Married" "Married"
 


# make a scatter plot with education on the x axis and wage on the y axis.



# recreate the previous plot, except add colors for male/female




# Use facet_wrap to break this scatterplot into seperate married and not married plots.




```


Why not make a simple histogram while we are here? 

```{r eval = FALSE}

```



## Updating Code: Control Structures

Control structures are not incredibly important in your use of R but they are typically important in coding
and we want to know what they are and how to use them. The nice thing about a script is we can plan out and edit an incredibly long list of complicated commands. Do this. Then that. Then that. Then that... This is great but sometimes we want to add some extra control to your commands. Maybe we want to do something 50 times. We'd like a way of commanding that. Or, maybe we have no idea exactly how many times we need something to done. Maybe we want R to do something until a condition is met. Or maybe we want R to do one thing in certain conditions and something else if not.

Control structures for, while, and if/else allow us to do precisely that. There are some efficiency issues with looping through these structures and we'll cover alternatives later.



```{r for_loop, eval=FALSE}

# check out the following code, modify it to print "one", "two", "three".

for (i in c(1:3)){
  print(i)
}

```

Use a for loop to calculate $10!$

```{r for_loop_problem}

# Modify the following for loop to calculate the factorial. We will set it to 1 at the beginning, after we exit your for loop we should be able to print out the factorial of 10.

factorial <- 1

for (i in 1:10){
factorial <- 1
}

print(factorial)
```


```{r while_loop, eval=FALSE}

# check out the following code, modify it such that we getting within 1 of solution is sufficient. You should probably change the output in the paste0 function to be clear we are within 1 of the solution.

solution <- sample(1:100, 1)
guess <- sample(1:100, 1)
counter <- 1

while (guess != solution){
  guess <- sample(1:100, 1)
  counter <- counter + 1
}

print(paste0("the solution is: ", guess, " and it only took us ", counter, " guesses."))

```


And finally, if...


```{r if_elseif_else}

# check out the following code, modify it such that it prints correct. Do not change the value of i. Change the if structure. There are endless ways to do this. Pick one.

i <- 50

if (i > 25){
  print("be more interesting")
} else if (i >20) {
  print("ummm?")
} else {
  print('correct')
}


```


### Here is something you can do if you want. I'm indifferent.

When we changed our guess loop to accept guesses within 1 of the solution, the expected number of guesses should decrease. Of course, expectations exist to be crushed in this one-time pass through existence, so maybe you found the solution in 1 guess with the stricter requirements and 40 guess with the less strict. Guess using the strict evaluation (guess = solution) 1000 times and calculate the average number of guesses. Do the same for the less strict version. Note that you probably want to get rid of the code that prints out the solution. You will need a while loop, you will need a for loop, and you will need a vector to keep track of how many guesses it took. You may or may not need other stuff.  

### Here is another thing you could do if you want. I don't care.

You see that even with a very terrible way of guessing a number, your computer can find it very quickly. I mean, it is possible to spend over 100 guesses trying to guess a number between 1 and 100. Nonetheless, computers do this stuff fast enough that the efficiency doesn't matter much at ths scale. But what if the problem was more complicated or the scale was larger. Try to improve our algorithm in two ways. First, do not repeat guesses. Once that is working, allow our guess to know whether we are too low or too high.

### Bogosort. I almost prefer you don't. 

R comes with efficient sort and order functions to arrange vectors and dataframes based on ascending or descending order. They work well. The absolute worst way to sort a vector is referred to as bogosort. Bogosort shuffles the vector and then checks if it was successful in sorting. If not, reshuffle. If you're really bored try using control structures to bogosort the sort_me vector in ascending order. First, try to use the sort() function to test your bogosort shuffle. If you get that to work, consider how you might do it without sort().

I can't stress how unimportant this is.


```{r bogosort, eval=FALSE}

sort_this <- rnorm(4)





```



## Including Math

There may be times where you want to use some math in your submission. This is very likely to be the case in your final paper. Even if it is not critically important, it is not difficult to do, so you should be aware it is an option. The syntax follows [LaTeX](https://en.wikibooks.org/wiki/LaTeX/Mathematics), although I believe there are some exceptions. Initially, you'll probably need to have the syntax guide open but it becomes intuitive quickly.

Just so you can see how it works. I've including a couple of examples. The following pairs will include math typed out in normal markdown to the best of my ability, followed by the LaTeX typesetted version. If you want to see what produced that version, look at the Rmd file.

* x = y + z

$$x = y + z$$

* x = (y/z)

$$x = \frac{y}{z}$$


* x^alpha + y^beta= z

$$x^\alpha + y^\beta= z$$


* same as above except subscripted as well

$$x^\alpha_i + y^\beta_i= z_i$$

* y-hat = ....

$$\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x_1 + \hat{\beta}_2 x_2$$



If you want to include the math inline you just use single dollar symbols. For example, you could include $e=s/v$ in this sentence. There is no need to practice this. If you need it, you just look up what symbols you need, but again, I want you to see it is painless enough to use if you want to.

We can also do R code inline. For example, here is a random number between 1 and 100: `r sample(1:100,1)`.



## 



